Finally Tagless Staged Interpreters for Simpler Typed Languages

### 2007-04-06 tagless/tagless-final.tex
Almost done with Section 5

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
state that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Roughly speaking, a partial evaluator interprets each object term as either
a static (present-stage) term (using the evaluator~|R|), or
a dynamic (future-stage) term, (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might try
to define partial evaluation via the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrInt :: P0 Int -> C Int
abstrInt (S0 r) = int (unR r)
abstrInt (E0 c) = c

abstrBool :: P0 Bool -> C Bool
abstrBool (S0 r) = bool (unR r)
abstrBool (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrInt e1)
                      (abstrInt e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (via~|R|) if and only if both operands
are static, otherwise we extract the dynamics terms from the operands and
add them using the compiler~|C|.  Uniformly, we use
the overloaded functions from the |Symantics|
instances for |R| and~|C| to define those for~|P0|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool| say) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}

@@

\section{Variations}\label{variations}

We show that our |Symantics| and generally our approach accomodates
a number of variants without any (new) difficulties.  In particular,
we show how a call-by-name CPS interpreter, a call-by-value CPS
program transformation.  Finally, we comment on the requirements on the type
system to support our approach.

@@

The object language generally inherits the evaluation strategy from
the metalanguage -- call-by-value (CBV) in OCaml, call-by-name (CBN) in Haskell.
One may wonder if it is
possible to represent a call-by-name object language in a
call-by-value metalanguage. The answer, given already by 
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}, is yes
thanks to CPS, which they specifically introduced 
to make the evaluation strategy of a definitional interpreter
independent of the strategy of the metalanguage. We demonstrate the
same independence in the typed setting. Specifically, we
demonstrate CBN evaluation for our object language in OCaml.
We do so by building a CBN CPS interpreter for the object
language.

The interpretation of an object term is a function
mapping a continuation~|k| to the answer
returned by~|k|.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = fun k ->
  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
\end{code}
In both cases, the returned value has type 
|(int -> 'w)->'w| where |'w| is the (polymorphic) answer type.

The interpretation of abstraction and application in CBN CPS is
different from CBV CPS and is as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> (f e2) k)
\end{code}
Charateristic of CBN, interpreting the application does not
``evaluate'' the argument |e2|, that is, we do apply it to the
continuation. Rather, we pass the unevaluated |e2| to the abstraction.
The result of |lam (fun x -> add x (int 1))| has the type
|((((int->'w1)->'w1) -> ((int->'w1)->'w1)) ->'w2)->'w2|.
We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, we encounter the same problem we had in
\S\ref{S:PE-problem}.
We cannot say that the result of interpreting an object
expression of type |t| is |(t->'w)->'w| because if |t| happens to
be an arrow type, the result
type obviously does not fit the desired pattern. We need a type
function with a typecase distinction.  Our previous solution 
(\S\ref{S:PE-solution}) where we use an extra type
argument |'sv| of the |repr| type, to encode such type-level functions,
also works here.  Furthermore, we notice that the type 
function |repr_pe| needed for the partial evaluator 
(in~\S\ref{S:PE-solution}) is precisely the same type function we
need for CBN CPS.

\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = 
    {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = 
    {ko = fun k -> e1.ko (fun v1 -> 
                   e2.ko (fun v2 -> k (v1+v2)))}
  let if_ eb et ee = 
    {ko = fun k -> eb.ko 
         (fun vb -> if vb then (et ()).ko k 
                          else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = 
    {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = 
    let rec fx f n = app (f (lam (fx f))) n in 
    lam (fx f)
  let get_res x = x.ko (fun v -> v)
end
\end{code}

We have made our interpreter fully polymorphic over the answer type.
We could have also made |RCN| a functor, parameterized over
the answer type -- as is common when writing CPS in SML. Here
we chose to use the higher-rank polymorphism offered by
OCaml record types.

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
under CBN but not CBV\@.
\begin{code}
module EXS(S: Symantics) = struct open S
 let diverg () = app (lam (fun x -> int 1)) 
                     (app (fix (fun f -> f)) (int 2))
end
\end{code}
If we interpret |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL}
\begin{code}
let module M = EXS(R) in M.diverg ()
\end{code}
we never get any result because the above diverges. In contrast, if we use
the CBN interpreter
\begin{code}
let module M = EXS(RCN) in RCN.get_res (M.diverg ())
\end{code}
we get the result |1|.

@@

We can easily change our CBN CPS interpreter into a CBV CPS
interpreter, by replacing only one definition:
\begin{code}
module RCV = struct
  include RCN
  let lam f = {ko = fun k -> k (fun e ->
    e.ko (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now when we apply an abstraction to an argument, we always force
the evaluation of that argument before proceeding. This approach is in
line with \citet{reynolds-relation} (albeit in a typed setting). The
interpreter RCV is useful for achieving CBV evaluation of the object language
whether the metalanguage is CBV or CBN.

In this section however we show a more general approach to CBV CPS:
a CPS transformer. We take any implementation of |Symantics| (interpreter,
compiler, partial evaluator) and make it evaluate the CPS language
instead. We can view interpreter transformation as a
transformation on the object language, from direct to
continuation-passing style.

The transformation is canonical, as the following (abbreviated) code
shows
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 -> S.app k (S.add v1 v2))))))
  ...
  let lam f = S.lam (fun k -> S.app k (S.lam (fun x ->
    f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v -> S.app (S.app f v) k)))))
  let fix f  = S.fix (fun self -> (f self))
end
\end{code}
The code explicitly establishes a mapping from CPS interpretations to
(direct) interpretations, as performed by 
the base interpreter |S|.

A reader may notice that the above module does not define |repr|
and thus does not have signature |Symantics|.
The reason for that is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has type\\
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we again need a 
type-discriminating
function similar to the type function |repr_pe| of
\S\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom) type variable.

@@

It is instructive to evaluate a few examples. With 
|CPST| instantiated with the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}

\noindent which shows the na\"{\i}ve CPS transformation of the original code.
The result is sub-optimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with |P| instead, we get the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
only together, never separately.  This treatment has the drawback that
we must duplicate a pre-encoded term in order to interpret it in
multiple ways.  The meta-notions of contexts and plugging may seem ad
hoc, but in fact they are just reflections of the type-class and module
machinery that we have been using in this paper all along.

@@

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be Jones\hyp optimal.
\begin{enumerate}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either using a
very advanced type system or incurring tagging/untagging overhead.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate}
While the logic is sound, we showed that the premise in the very first step
was not valid.

@@

A lot of effort has gone into finding ways to give precise types to untyped
terms (see for example \citet{baars-typing,Guillemette-Monier-PLPV}
as well as the messages \citet{haskell-list}).  These 
take advantage of the host language's types to varying extents.  While
very successful for proper terms, results for
ill-typed terms can produce rather obscure error messages.  We use the
host's typing facilities for this, and thus inherit the strengths
and weaknesses of the host language.  While that may not be optimal, it is
certainly much more modular.

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on coalgebraic structures in the
$\lambda$\hyp calculus

>>>>

Our contributions are as follows.
\begin{enumerate*}
\item We build interpreters that evaluate (\S\ref{language}),
    compile (or evaluate with staging) (\S\ref{compiler}), and partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing styles
   (\S\ref{variations}).
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our interpreters, from
    evaluation and compilation (\S\ref{language}) to partial evaluation and CPS transformation (\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state (\S\ref{state}).
\item We describe an approach to self\hyp interpretation compatible with the
  above (\S\ref{selfinterp}).  Self\hyp interpretation turned out harder than expected.
\end{enumerate*}
Our code is surprisingly simple and obvious in hindsight, but
it has been an open problem to
interpret a typed object language in a typed metalanguage without
tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

Roughly, a partial evaluator interprets each object term to either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, one might define
the Haskell data type
\begin{code}
data P0 t = S0 (R t) | E0 (C t)
\end{code}
To extract a dynamic term from this type, we create the functions
\begin{code}
abstrI :: P0 Int -> C Int
abstrI (S0 r) = int (unR r)
abstrI (E0 c) = c

abstrB :: P0 Bool -> C Bool
abstrB (S0 r) = bool (unR r)
abstrB (E0 c) = c
\end{code}
We then start to define |P0| as an instance of the |Symantics| class.
\begin{code}
instance Symantics P0 where
  int  x = S0 (int x)
  bool x = S0 (bool x)
  add (S0 e1) (S0 e2) = S0 (add e1 e2)
  add e1 e2 = E0 (add (abstrI e1) (abstrI e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R|) if and only if both operands
are static; otherwise we extract the dynamic terms from the operands and
add them using~|C|.  Thus the two uses of |add| above refer to
|add| for~|R| and |add| for~|C|.

Whereas |mul| and |leq| are as easy to define as |add|, we encounter
a problem with |if_|.  If the first argument to |if_| is a dynamic term
(type |C Bool|) and the second and third arguments are a static term
(type |R a|) and a dynamic term (type |C a|), then we need to convert
the static term to dynamic, but there is no polymorphic ``lift''
function, of type |a -> C a|, to send a value to the future stage
\citep{xi-guarded,WalidPOPL03}.\footnote{We note in passing that, if we
were to add polymorphic \texttt{lift} to the type class
\texttt{Symantics repr}, then \texttt{repr} would become an instance of
\texttt{Applicative} and thus \texttt{Functor}:\quad\texttt{fmap
f = app (lift f)}}

@@

\section{Continuation\hyp passing style}\label{variations}

Our approach accommodates
several variants, including
a call-by-name CPS interpreter and a call-by-value CPS
transformation.

@@

The object language generally inherits the evaluation strategy from
the metalanguage---call-by-value (CBV) in OCaml, call-by-name (CBN) in
Haskell.  To represent a CBN object language in a CBV metalanguage,
\citet{reynolds-definitional,reynolds-relation} and \citet{PlotkinCBN}
introduce CPS to make the evaluation strategy of a definitional
interpreter indifferent to that of the metalanguage. To achieve the same
indifference in the typed setting, we build a CBN CPS interpreter for
our object language in OCaml.

The interpretation of an object term is a function
mapping a continuation~|k| to the answer
returned by~|k|.
\begin{code}
let int (x:int) = fun k -> k x
let add e1 e2 = fun k ->
  e1 (fun v1 -> e2 (fun v2 -> k (v1 + v2)))
\end{code}
In both |int| and |add|, the interpretation has type 
|(int -> 'w)|\texttt{ }|-> 'w|, where |'w| is the (polymorphic) answer type.

Unlike CBV CPS, the CBN CPS interprets
abstraction and application as follows:
\begin{code}
let lam f = fun k -> k f
let app e1 e2 = fun k -> e1 (fun f -> f e2 k)
\end{code}
Charateristic of CBN, |app e1 e2|
does not evaluate the argument~|e2| by applying it to the
continuation~|k|. Rather, it passes |e2| unevaluated to the abstraction.
Interpreting $\fun{x} x+1$ yields type
\begin{code}
((((int -> 'w1) -> 'w1) -> (int -> 'w1) -> 'w1)
 -> 'w2) -> 'w2
\end{code}

We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
an object term inductively determines the type of its interpretation:
the interpretation of an object term of type~$t$ may not have type
|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
simulate a type function with a typecase distinction, using an extra
type argument to |repr|. Happily, the type function |repr_pe| needed for
the partial evaluator 
in~\S\ref{S:PE-solution} is precisely the same type function we
need for CBN CPS\@.
\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = 
    {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = 
    {ko = fun k -> e1.ko (fun v1 -> 
                   e2.ko (fun v2 -> k (v1+v2)))}
  let if_ eb et ee = 
    {ko = fun k -> eb.ko 
         (fun vb -> if vb then (et ()).ko k 
                          else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = 
    {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = 
    let rec fx f n = app (f (lam (fx f))) n in 
    lam (fx f)
  let run x = x.ko (fun v -> v)
end
\end{code}

This interpreter~|RCN| is fully polymorphic over the answer type,
using higher-rank polymorphism through OCaml record types.
It could also be a functor parameterized over
the answer type.

Because |RCN| has the signature |Symantics|, we can instantiate our previous
examples with it, and all works as expected.  More interesting
is the example $(\fun{x}1)\bigl((\fix{f}f)\mathinner2\bigr)$, which terminates
under CBN but not CBV\@.
\begin{code}
module EXS(S: Symantics) = struct open S
 let diverg () = app (lam (fun x -> int 1)) 
                  (app (fix (fun f->f)) (int 2))
end
\end{code}
Interpreting |EXS| with the |R| interpreter of
\S\ref{S:interpreter-RL} does not terminate.
\begin{code}
let module M = EXS(R) in M.diverg ()
\end{code}
In contrast, the CBN interpreter gives the result~|1|.
\begin{code}
let module M = EXS(RCN) in RCN.run (M.diverg ())
\end{code}

@@

Changing one definition turns our CBN CPS interpreter into CBV\@.
\begin{code}
module RCV = struct include RCN
  let lam f = {ko = fun k -> k (fun e ->
    e.ko (fun v -> f {ko = fun k -> k v}))}
end
\end{code}
Now an applied abstraction
evaluates its argument before proceeding. This approach is in
line with \citets{reynolds-relation}, albeit typed. The
interpreter~|RCV| is useful for CBV evaluation of the object language
whether the metalanguage is CBV or CBN\@.

We turn to a more general approach to CBV CPS: a CPS transformer that
turns any implementation of |Symantics| into a CPS interpreter, whether
it is an evaluator.  This functor on interpreters performs a textbook
CPS transformation on the object language.
\begin{code}
module CPST(S: Symantics) = struct
  let int i = S.lam (fun k -> S.app k (S.int i))
  let add e1 e2 = S.lam (fun k ->
    S.app e1 (S.lam (fun v1 ->
    S.app e2 (S.lam (fun v2 ->
    S.app k (S.add v1 v2))))))
  let lam f = S.lam (fun k -> S.app k (S.lam
    (fun x -> f (S.lam (fun k -> S.app k x)))))
  let app e1 e2 = S.lam (fun k -> 
    S.app e1 (S.lam (fun f ->
    S.app e2 (S.lam (fun v ->
    S.app (S.app f v) k)))))
  let fix f = S.fix (fun self -> (f self))
end
\end{code}
The (abbreviated) code explicitly maps CPS interpretations to
(direct) interpretations performed by 
the base interpreter~|S|.

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the result type of |lam f|. Whereas
|int i| or |add e1 e2| have the (abbreviated) 
type |('c,..., (int -> 'w) -> 'w) S.repr|,
the result of |lam (fun x -> add x (int 1))| has type\\
|('c,...((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr|. 
Hence, to write the type equation for |CPST.repr| we again need a 
type-discriminating
function similar to the type function |repr_pe| of
\S\ref{S:PE-solution}. Alas, the function we need here is
similar but not identical to |repr_pe|. Thus we need to add 
another type variable to the |repr| type and modify the Symantics
signature accodringly. Again, the terms in previous Symantics module
stay unchanged, but the |repr| type equations in those structures have to
account for a new (phantom) type variable.

@@

It is instructive to evaluate a few examples. With 
|CPST| instantiated with the compiler~|C| above,
|T.test1| gives
\begin{code}
.<fun x_5 -> ((fun x_2 -> 
        (x_2 (fun x_3 -> fun x_4 -> (x_4 x_3))))
        (fun x_6 -> ((fun x_1 -> (x_1 (true))) 
        (fun x_7 -> ((x_6 x_7) x_5)))))>.
\end{code}
which shows the na\"{\i}ve CPS transformation of the original code.
The result is sub-optimal with several apparent $\beta$-redexes. 
By instantiating |CPST| with |P| instead, we get the expected
\begin{code}
  P.repr = {P.st = Some <fun>;
            P.dy = .<fun x_5 -> (x_5 (true))>.}
\end{code}

@@

We type-check a pre-encoded term~$\encode{e}$ and its interpreter~$I[~]$
only together, never separately.  This treatment has the drawback that
we must duplicate a pre-encoded term in order to interpret it in
multiple ways.  The meta-notions of contexts and plugging may seem ad
hoc, but in fact they just reflect the type-class and module
machinery that we have been using in this paper all along.

@@

\Citet{WalidICFP02}, \citet{taha-tag}, \citet{xi-guarded}, and
\citet{peyton-jones-simple} seem to argue as follows that a self\hyp
interpreter of a typed language cannot be Jones\hyp optimal.
\begin{enumerate*}
\item One needs to encode a typed language in a typed language based on
a sum type (at some level of the hierarchy),
\item It is not possible to write a \emph{direct} interpreter 
for such an encoding of a typed language
in a typed language, without either an
advanced type system or the overhead of tagging and untagging.
\item Thus an indirect interpreter is necessary, which needs a universal
  type (and hence tagging),
\item Thus any self-interpreter must have tags.
\item Hence the self-interpreter cannot be Jones-optimal.
\end{enumerate*}
While the logic is sound, we showed that the premise in the very first step
was not valid.

@@

A lot of effort has gone into finding ways to give precise types to untyped
terms \citep{baars-typing,Guillemette-Monier-PLPV,haskell-list}).  These 
take advantage of the host language's types to varying extents.  While
very successful for proper terms, results for
ill-typed terms can produce rather obscure error messages.  We use the
host's typing facilities for this, and thus inherit the strengths
and weaknesses of the host language.  While that may not be optimal, it is
certainly much more modular.

@@

\acks%\subsection*{Acknowledgments}
We thank Martin Sulzmann and Walid Taha 
for helpful discussions.  Sam Staton, Pieter Hofstra, and Bart Jacobs
kindly provided some useful references on coalgebraic structures in the
$\lambda$\hyp calculus.
